Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2020-06-30 pkg/circtree/inst/wind_paper/draft/draft.Rnw
Touchups and put Equation in brackets
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3023 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a8d1b637a940e6959d8d30df4a83a3805ff774b9

The score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
provides a way to obtain a measure of goodness of fit of the model for each
observation and fitted parameter. Then, the optimization problem in
Equation~\ref{eq:optim} can alternatively be specified as 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
Figure~\ref{fig:densities} depicts a von~Mises model for circular data in $[0, 2
\pi)$ fitted by maximum likelihood, either using a linearized (left) or
circular (right) scale. In both cases, the empirical histogram (gray bars) is
shown along with the fitted density (red line) and estimated location parameter
(red hand). However, this distributional model only considers the circular
response variable but no covariate. Of course, including covariates is of
interest in a regression setup for forecasting.

@@

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with substantial variations. Therefore, separating the data
set into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (potential) group-specific
effects more precisely and hence can result in an overall better-fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019}. Specifying a
full distributional model in each node of the tree yields a distributional
regression tree, where selecting the von~Mises distribution enables an
application to circular data. The crucial step of how and where to split the
data can be accomplished with the unbiased recursive partitioning algorithms
MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
obtained by evaluating the score function~$s(\cdot)$ for each individual
observation at the parameter estimates (Equation~\ref{eq:scores}). For the
von~Mises distribution with its two distribution parameters ($\mu$ and
$\kappa$) and a data set of $n$ observations, this yields an $n \times 2$
matrix that can be employed as a discrepancy measure, capturing how well each
given observation conforms with the estimated location $\hat{\mu}$ and
precision $\hat{\kappa}$, respectively. To capture dependence on covariates,
the association between the model's scores and each available covariate is
assessed using either a parameter instability test~(MOB) or a permutation
test~(CTree). By doing so in each partitioning step, the covariate with the
highest significant association (i.e., lowest significant $p$-value, if any) is
selected for splitting the data. The corresponding split point is chosen either
by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions. This procedure is repeated recursively until
there are no significant parameter instabilities or until another stopping
criterion is met (e.g., subgroup size or tree depth). A more detailed
description of the applied tree-building algorithm can be found in
Appendix~\ref{app:algorithm}.

@@

    implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular covariates would require
    an additional manual adjustment, such as an approximation by a finite Fourier expansion
    (cf., \citealt{Lund:1999}), we only use the linear components of the lagged
    2-dimensional wind vector~$(u, v)^{\top}$ and the lagged wind speed $spd$ as
    predictor variables. Thus, the model formula for the location
    parameter $\mu$ of the von~Mises distribution can be written as:

@@

The performance of the novel circular regression trees and forests is assessed
in an application on shortterm probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In
summary, the circular regression trees and forests have the highest predictive
performance in this setting. For cases without changes in the wind regime,
lagged response values provide already highly skillful estimates leading to a
good performance of the persistency model as observed for shortterm wind
direction forecasts in the application. While in these cases the trees and forests
also benefit from the highly informative lagged response, they can compensate
for a lower information of this covariate by incorporating other quantities and
possible interactions of these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods.

>>>>

The score function
%%
\begin{equation}
  \label{eq:scores}
  \begin{aligned}
    s(\mu, \kappa, y) &= \bigg(\frac{\partial \ell}{\partial \mu}(\mu,\kappa; y)
    \smallskip &,
    \quad & \frac{\partial \ell}{\partial \kappa}(\mu,\kappa; y) \bigg) 
    \\
    &= \bigg(\kappa \sin(y-\mu)
    \smallskip &,
    \quad &-\frac{I_1(\kappa)}{2\pi I_0(\kappa)}+\cos(y-\mu) \bigg)
  \end{aligned}
\end{equation}
%%
provides a way to obtain a measure of goodness of fit of the model for each
observation and fitted parameter. Then, the optimization problem in
Equation~(\ref{eq:optim}) can alternatively be specified as 
%%
\begin{equation}
  \sum_{i = 1}^n s(\hat{\mu}, \hat{\kappa}, y_i) = 0.
\end{equation}
%%
Figure~\ref{fig:densities} depicts a von~Mises model for circular data in $[0,
2 \pi)$ fitted by maximum likelihood, either using a linearized~(left) or
circular~(right) scale. In both cases, the empirical histogram (gray bars) is
shown along with the fitted density (red line) and estimated location
parameter~(red hand). However, this distributional model only considers the
circular response variable but no covariate. Of course, including covariates is
of interest in a regression setup for forecasting.

@@

\subsection{Circular regression trees}
\label{sec:circtree}
Fitting a global model to a full data set can be very challenging, particularly
for complex data with substantial variations. Therefore, separating the data
set into more homogeneous subgroups based on covariates before fitting a local
model in each of these subgroups allows to capture (potential) group-specific
effects more precisely and hence can result in an overall better-fitting model.
This is the general idea of regression trees which are combined with
distributional modeling in \cite{Schlosser+Hothorn+Stauffer:2019}. Specifying a
full distributional model in each node of the tree yields a distributional
regression tree, where selecting the von~Mises distribution enables an
application to circular data. The crucial step of how and where to split the
data can be accomplished with the unbiased recursive partitioning algorithms
MOB \citep{Zeileis+Hothorn+Hornik:2008} or CTree
\citep{Hothorn+Hornik+Zeileis:2006}. For this purpose, model scores are
obtained by evaluating the score function~$s(\cdot)$ for each individual
observation at the parameter estimates (Equation~(\ref{eq:scores})). For the
von~Mises distribution with its two distribution parameters ($\mu$ and
$\kappa$) and a data set of $n$ observations, this yields an $n \times 2$
matrix that can be employed as a discrepancy measure, capturing how well each
given observation conforms with the estimated location $\hat{\mu}$ and
precision $\hat{\kappa}$, respectively. To capture dependence on covariates,
the association between the model's scores and each available covariate is
assessed using either a parameter instability test~(MOB) or a permutation
test~(CTree). By doing so in each partitioning step, the covariate with the
highest significant association (i.e., lowest significant $p$-value, if any) is
selected for splitting the data. The corresponding split point is chosen either
by optimizing the log-likelihood~(MOB) or a two-sample test statistic~(CTree)
over all possible partitions. This procedure is repeated recursively until
there are no significant parameter instabilities or until another stopping
criterion is met (e.g., subgroup size or tree depth). A more detailed
description of the applied tree-building algorithm can be found in
Appendix~\ref{app:algorithm}.

@@

    implementation of \cite{Mulder+Klugkist:2017} the accommodation of circular
    covariates would require an additional manual adjustment, such as an
    approximation by a finite Fourier expansion (cf., \citealt{Lund:1999}), we only
    use the linear components of the lagged 2-dimensional wind vector~$(u,
    v)^{\top}$ and the lagged wind speed $spd$ as predictor variables. Thus, the
    model formula for the location parameter $\mu$ of the von~Mises distribution
    can be written as:

@@

The performance of the novel circular regression trees and forests is assessed
in an application on short-term probabilistic wind direction forecasting at two
airports with different site characteristics. As benchmark models,
probabilistic climatology and persistency models, as well as a state-of-the-art
circular GLM-type model are evaluated based on proper scoring rules. In
summary, the circular regression trees and forests have the highest predictive
performance in this setting. For cases without changes in the wind regime,
lagged response values provide already highly skillful estimates leading to a
good performance of the persistency model as observed for short-term wind
direction forecasts in the application. While in these cases the trees and forests
also benefit from the highly informative lagged response, they can compensate
for a lower information of this covariate by incorporating other quantities and
possible interactions of these, contrary to the persistency model (see
Figure~\ref{fig:boxplot_crpsskill}). Hence, the tree-based models provide
reliable forecasts in all tested meteorological settings. For operational use,
a possible extension could be the incorporation of numerical weather
predictions as (additional) covariates. While this probably only slightly
improves the predictive skill for short leadtimes, it possibly extends the
potential forecast range of the different methods.
